lme4ord (l-m-e-ford)
====================

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(fig.path = "inst/README/figure/")
```

Mixed-effects models for community ecologists.  See the currently
evolving [mission statement](https://github.com/stevencarlislewalker/lme4ord/issues/1).

This package is not at all stable.

```{r, message = FALSE}
library(Matrix)
library(lme4ord)
library(plotrix)
library(minqa)
library(ape)
library(lme4)
library(lme4pureR)
library(multitable)
library(pryr)
library(reo)
```

#### phylogenetic generalized linear mixed models!

Begin with initial simulations of a sites-by-species binary response
matrix, `y`, environmental variable, `x`, and trait `z`.
stronger correlations between `y` and `x` will be added below.
```{r}
set.seed(10)
n <- 10
m <- 30
dl <- dims_to_vars(data.list(y = 1 * (rmat(n, m) > 0),
                             x = rnorm(n), z = rnorm(m),
                             dimids = c("sites", "species")))
df <- as.data.frame(dl)
head(df)
```

Make up some silly phylogeny.
```{r}
phy <- rtree(n = m)
phy <- compute.brlen(phy, method = "Grafen", power = 0.5)
```
and estimate a phylogenetic covariance matrix, standardized to unit determinant.
```{r}
Vphy <- stanCov(vcv(phy))
dimnames(Vphy) <- rep(list(1:m), 2)
```
Here's the phylogeny (forget the species names) and the associated covariance matrix
```{r, fig.width = 4, fig.height = 5}
plot(phy)
image(as(Vphy, "sparseMatrix"))
```
Put the covariance matrix in a list, for model-input purposes -- the
idea is that there might be other covariance matrix (e.g. a spatial
one say).  It is important that the list element gets the name
`species` because this is the name of the grouping factor used in the
model formula below.
```{r}
covList <- list(species = Vphy)
```

Here is the cool part ... a formula interface.  This model has a fixed
interaction between the environment and the trait (with intercept and
main effects too), a random environmental slope and intercept with
phylogenetic correlations across species.  However, the phylogenetic
nature of the covariances is not set in the formula, but rather as an
argument to the `glmercFormula` function below, which will form the
formula parsing module of a pglmer function.
```{r}
form <- y ~ x*z + (x | species)
parsedForm <- glmercFormula(form, df, covList = covList)
```

Set the covariance parameters to something more interesting (i.e. with
a covariance between the slope and intercept).
```{r}
covarSim <- c(0.5, -0.2, 0.5)
parsedForm <- within(parsedForm, Lambdat@x[] <- mapToCovFact(covarSim))
```
Update the simulations to reflect the new structure.
```{r}
X <- model.matrix(nobars(form), df) # fixed effects design matrix
Z <- t(parsedForm$Lambdat %*% parsedForm$Zt) # random effects design
                                             # matrix with
                                             # phylogenetic
                                             # covariances
fixefSim <- rnorm(ncol(X)) # fixed effects
u <- rnorm(ncol(Z)) # whitened random effects
p <- plogis(as.numeric(X %*% fixefSim + Z %*% u)) # probability of observation
dl$y <- rbinom(nrow(df), 1, p) # presence-absence data
df <- as.data.frame(dl) # reconstruct the data frame with new
                        # structured response
```
Now we look at the new structure.  Here's the Cholesky factor of the species covariance, and the covariance itself.
```{r, fig.width=3, fig.height=3}
image(parsedForm$Lambdat)
image(crossprod(parsedForm$Lambdat))
```
The big four blocks represent the 2-by-2 covariance between intercept
and slope.  The covariances within these blocks represent phylogenetic
covariance.  the pattern here is more closely related species have
more similar intercepts and slopes (red blocks on the diagonal) but
more closely related species also have stronger negative correlations
between slope and intercept (blue blocks on off diagonal).

Here's the transposed random effects model matrix.  Those are 1's for
the intercepts in the first 30 rows and the environmental variable in
the second 30.
```{r, fig.width=7, fig.height=2}
image(parsedForm$Zt)
```

Here's the full covariance matrix (the large scale blocks reflect
phylogenetic correlations and the patterns within each block are due
to the environmental variable).
```{r, fig.width=5, fig.height=5}
image(fullCov <- t(parsedForm$Zt) %*% crossprod(parsedForm$Lambdat) %*% parsedForm$Zt)
```
Here's a closeup of one of the blocks
```{r, fig.width=3, fig.height=3}
image(fullCov[1:10, 1:10])
```
A potential problem is that this block is singular.
```{r}
eigen(fullCov[1:10, 1:10])$values
```
In fact the rank of the full 300 by 300 matrix is only 60 = 30 species
times 2 model matrix columns.
```{r}
rankMatrix(fullCov)[1]
```
But then again so is the standard non-phylogenetic `glmer` model.
```{r}
gm <- glmer(form, df, binomial)
with(getME(gm, c("Zt", "Lambdat")), {
    covMatGm <- t(Zt) %*% crossprod(Lambdat) %*% Zt
    print(rankMatrix(covMatGm)[1])
    dim(covMatGm)
})
```
The distribution of underlying probabilities of occurrence looks OK.
```{r, fig.width=3, fig.height=3}
hist(p)
```
Here is the observed occurrence pattern.
```{r, fig.width=3, fig.height=3}
color2D.matplot(dl$y, xlab = "species", ylab = "sites", main = "abundance")
```

```{r}
mod <- glmerc(form, df, covMat = covMat)
```
and compare with the true parameter values.
```{r}
cbind(estimated = mod$opt$par, # estimated parameters
      true = c(covar = covarSim, fixef = fixefSim)) # true parameters
```
Looks great!  At least in this case.

#### mixed effects ordination!

```{r}
data(fish)
data(limn)
Y <- as.matrix(fish)
## Y <- Y[, colSums(Y) > 1]
n <- nrow(Y)
m <- ncol(Y)
x <- as.vector(scale(limn$pH))
dl <- data.list(Y = t(Y), x = x,
                dimids = c("species", "sites"))
summary(dl)
```

Not done!