Notes on `glmer` deviance functions
===================================

```{r}
library("lme4")
```

```{r} 
parsedForm <- glFormula(cbind(incidence, size - incidence) ~ period + (1 | herd),
                        family = binomial, data = cbpp)
```

`glmer` models are usually fitted in two steps.  The first step
involves no integration (`nAGQ = 0`), and the second involves
integration by either Laplace (`nAGQ = 1`) or adaptive Gaussian
quadrature (`nAGQ > 1`).  

```{r}
(devfun0 <- do.call(mkGlmerDevfun, parsedForm))
```

```{r}
(devfun1 <- updateGlmerDevfun(devfun0, parsedForm$reTrms))
```

```{r}
as.list(body(devfun1)[c(3, 4, 8, 9)])
```

#### First stage deviance function

Note that there are several objects being used that do not appear to
be created anywhere (e.g. `lp0`).  Don't worry, they are actually in the
environment of `devfun0`, which we can list using the following
command.
```{r}
ls(rho <- environment(devfun0))
```
Therefore, all of these objects are available to `devfun0`.

We will now step through each line, explaining what it does.  The
first line,
```{r, echo=FALSE}
body(devfun0)[[2]]
```
updates the conditional mean of the response variable at the value of
the linear predictor, `lp0`.  Mathematically, the linear predictor is,
$$ \mathbf\eta = \mathbf o + \mathbf X\mathbf\beta + \mathbf Z
\mathbf\Lambda_\theta \mathbf u $$
The conditional mean is then,
$$ \mathbf\mu = \text{link}(\mathbf\eta) $$
Note that the condition mean depends on the covariance parameters
$\mathbf\theta$.  Therefore, this update of the conditional mean is
based on the old value of $\mathbf\theta$.  The new value of
$\mathbf\theta$ is updated on the next line:
```{r, echo=FALSE}
body(devfun0)[[3]]
```
Next the penalized weighted residual sum of squares is calculated with
this new value of $\mathbf\theta$:
```{r, echo=FALSE}
body(devfun0)[[4]]
```
Finally the weights are updated (for safety?)
```{r, echo=FALSE}
body(devfun0)[[5]]
```


------------------------------------------------------------

#### Second stage deviance function



```{r}
ls(rho <- environment(devfun1))
```

```{r}
rho$baseOffset
```


```{r, echo=FALSE}
body(devfun1)[[2]]
```

```{r, echo=FALSE}
body(devfun1)[[3]]
```

```{r, echo=FALSE}
body(devfun1)[[4]]
```

```{r, echo=FALSE}
body(devfun1)[[5]]
```

```{r, echo=FALSE}
body(devfun1)[[6]]
```

```{r, echo=FALSE}
body(devfun1)[[7]]
```



